### 2009-11-11 docs/marcello-dissertation/thesis/chapter1.tex
* This revision fixes the problem related to the table of contents that was not printed.   - The package for hiper-link is responsible for the bug in the preamble.tex.
* References to the misc should be done by using \ref instead of \cite.

* Renaming a few bibliographic entries.

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@635 559edde2-a4d4-44e0-96fc-4035e40d7396

Sensor networks are commonly used in different areas for different purposes
such as scientific measurements and general population use
\cite{snintro01} \cite{snintro02}. In the scientific community, they serve as
tools to monitor the state of the environment by storing samples of data in specific 
periods of time. For instance, the NASA Jet Propulsion Laboratory uses the 
Volcano SensorWeb \cite{snex02}, a sensor network which aims to collect data
from specific volcanoes in Alaska and South Pacific to be used with heuristics 
regarding dangerous activities. Similarly, the National Data Buoy Center, a
department of the National Oceanic and Atmospheric Administration (NOAA)
\cite{snex03}, provides online data from different buoys in different coast
shores around the world regarding water quality, temperature, tides, etc, what 
characterizes sensor network for environmental monitoring \cite{snex01}.
Therefore, sensor networks can be seen as a very important scientific research and 
commercial field with direct application and use of its produced data.

@@

When it comes to data model implementation, the relational data model
\cite{db01} is one of the most used approaches to represent data in sensor 
network databases in any of the query processing strategies, and in most cases 
with a modified version of the SQL language \cite{sn-db-newop01}
\cite{sn-db-newop02}. However, the design of the database system must be done
with prior knowledge of the sensor types when  choosing the relational data
model, as the database design must be normalized \cite{db02} for the entities
already identified for the system. For instance, the introduction of a new
sensor device might represent a potential change on the database structure by
seeking a new normalized version of the database structure. In addition to
challenges of maintaining the database schema normalized, the use of Data
Provenance approaches in sensor network data \cite{sn-provenance01} usually
addresses the problem related with the lineage of data  and its full history.
However, once the data model is in place to receive the collected data, one of
the most important non-functional requirements in sensor networks is
scalability of the persistence storage. Since an operational sensor network
can produce huge amounts of data, it must be able to cope with the increase
and data load during data gathering, and yet provide users the same
performance during data retrieval by using different strategies such a
data-centric storage \cite{sn-storage03}.

In summary, this dissertation discusses the solution of a centralized
data-centric persistence storage component for the NetBEAMS \cite{nb01}, Data
Sensor Platform (DSP), an open-source \cite{se01} implementation of a sensor
network that primarily focused on environmental data produced by NetBEAMS. 
Different challenges can be related to the DSP in the context of sensor 
networks: data model, data persistence organization and strategy, and user data
access and distribution. First, NetBEAMS needs to provide means to safely 
represent sensor devices' properties without the need of restructuring any 
existing data model, taking into account the use cases from the main users. 
For example, a marine biologist may query the the persistence storage regarding
the temperature of a specific geographic region at the San Francisco Bay at a 
given time frame. Second, since a sensor network may experiment grown in terms 
of number of sensors, and consequently the number of data produced, the
persistence  storage must be able to reasonably scale and be well structured
in terms of storage space with the least service interruption. For instance,
if the persistence storage is close to reach its storage capacity, what
should the network administrator do? Last, but not least, the means to access 
the data for reuse is the most important features of the system, taking into 
account the different data formats used by the main users of the system. How 
to export specific collected data to a spreadsheet or archive files to be 
published or shared with other users? For these, and many other reasons, the 
persistence storage problem addressed by this work can be seen as a very 
challenging experience to solve.

In view of the fact that the persistence storage of a sensor network imposes
many different challenges, and taking into account the case study NetBEAMS, this
work's main strategy focused analyzing the different specifications of data
persistence in the literature in order to find new types of mechanisms to
provide data persistence for sensor network data. In this way, the kernel of
this dissertation is focused on providing solutions to the real-world problem
face by the Romberg Tiburon Center and its use of NetBEAMS. Furthermore,
the evaluation of different technologies is discussed in order to provide a
loosely-coupled component-based solution based on the infrastructure of the 
DSP Platform. For the reason of providing a 'lighter' data model, a subtype of
the Key-Value Pair \cite{kvp01} data model was chosen to easily capture the
description of any sensor device's properties, along with the use of time
series data models \cite{sn-provenance01} for time snapshots. In that way, data
modeling process does not represent a new refactoring each time a new sensor 
type is added or removed from the persistence system, giving users means to 
retrieve data accordingly. Similarly, in contemplation of scalability, this 
work proposes the use of an increasingly popular infrastructure technique called
Database Sharding \cite{dbshard01} \cite{dbshard02}, which takes the approach of
Data-Centric storage \cite{sn-storage03} on in-network query processing in a
centralized database system. In order to provide such mechanism, this work
evaluates the use of shard keys based on the geographic origin of the collected
data as cited to be one of the options when working with Provenance data in 
sensor networks \cite{sn-provenance01}. As a result, the data persistence system
component was conceived using the plug-and-play capability from the DSP 
platform without doing  any changes to the current DSP Platform, providing an
scalable solution that can adjust according to any data load increase. Finally,
the selected technology provides different data access mechanisms, which
directly addressed problems related to data access and Information Fusion 
\cite{sndp06}, providing export-import capabilities and visualization through
the availability of APIs written in different popular programming languages.

>>>>

Sensor networks are commonly used in different areas for different purposes
such as scientific measurements and general population use
\cite{sn-intro01} \cite{sn-intro02}. In the scientific community, they serve as
tools to monitor the state of the environment by storing samples of data in specific 
periods of time. For instance, the NASA Jet Propulsion Laboratory uses the 
Volcano SensorWeb \cite{sn-ex02}, a sensor network which aims to collect data
from specific volcanoes in Alaska and South Pacific to be used with heuristics 
regarding dangerous activities. Similarly, the National Data Buoy Center, a
department of the National Oceanic and Atmospheric Administration (NOAA)
\cite{sn-ex03}, provides online data from different buoys in different coast
shores around the world regarding water quality, temperature, tides, etc, what 
characterizes sensor network for environmental monitoring \cite{sn-ex01}.
Therefore, sensor networks can be seen as a very important scientific research and 
commercial field with direct application and use of its produced data.

@@

When it comes to data model implementation, the relational data model
\cite{relational-model} is one of the most used approaches to represent data in
sensor network databases in any of the query processing strategies, and in most cases 
with a modified version of the SQL language \cite{sn-db-newop01}
\cite{sn-db-newop02}. However, the design of the database system must be done
with prior knowledge of the sensor types when  choosing the relational data
model, as the database design must be normalized \cite{db-normalization} for the entities
already identified for the system. For instance, the introduction of a new
sensor device might represent a potential change on the database structure by
seeking a new normalized version of the database structure. In addition to
challenges of maintaining the database schema normalized, the use of Data
Provenance approaches in sensor network data \cite{sn-provenance01} usually
addresses the problem related with the lineage of data  and its full history.
However, once the data model is in place to receive the collected data, one of
the most important non-functional requirements in sensor networks is
scalability of the persistence storage. Since an operational sensor network
can produce huge amounts of data, it must be able to cope with the increase
and data load during data gathering, and yet provide users the same
performance during data retrieval by using different strategies such a
data-centric storage \cite{sn-storage03}.

In summary, this dissertation discusses the solution of a centralized
data-centric persistence storage component for the NetBEAMS \cite{netbeams09},
Data Sensor Platform (DSP), an open-source \cite{open-source} implementation of
a sensor network that primarily focused on environmental data produced by NetBEAMS. 
Different challenges can be related to the DSP in the context of sensor 
networks: data model, data persistence organization and strategy, and user data
access and distribution. First, NetBEAMS needs to provide means to safely 
represent sensor devices' properties without the need of restructuring any 
existing data model, taking into account the use cases from the main users. 
For example, a marine biologist may query the the persistence storage regarding
the temperature of a specific geographic region at the San Francisco Bay at a 
given time frame. Second, since a sensor network may experiment grown in terms 
of number of sensors, and consequently the number of data produced, the
persistence  storage must be able to reasonably scale and be well structured
in terms of storage space with the least service interruption. For instance,
if the persistence storage is close to reach its storage capacity, what
should the network administrator do? Last, but not least, the means to access 
the data for reuse is the most important features of the system, taking into 
account the different data formats used by the main users of the system. How 
to export specific collected data to a spreadsheet or archive files to be 
published or shared with other users? For these, and many other reasons, the 
persistence storage problem addressed by this work can be seen as a very 
challenging experience to solve.

In view of the fact that the persistence storage of a sensor network imposes
many different challenges, and taking into account the case study NetBEAMS, this
work's main strategy focused analyzing the different specifications of data
persistence in the literature in order to find new types of mechanisms to
provide data persistence for sensor network data. In this way, the kernel of
this dissertation is focused on providing solutions to the real-world problem
face by the Romberg Tiburon Center and its use of NetBEAMS. Furthermore,
the evaluation of different technologies is discussed in order to provide a
loosely-coupled component-based solution based on the infrastructure of the 
DSP Platform. For the reason of providing a 'lighter' data model, a subtype of
the Key-Value Pair \cite{db-kvp01} data model was chosen to easily capture the
description of any sensor device's properties, along with the use of time
series data models \cite{sn-provenance01} for time snapshots. In that way, data
modeling process does not represent a new refactoring each time a new sensor 
type is added or removed from the persistence system, giving users means to 
retrieve data accordingly. Similarly, in contemplation of scalability, this 
work proposes the use of an increasingly popular infrastructure technique called
Database Sharding \cite{db-shard01} \cite{db-shard02}, which takes the approach
of Data-Centric storage \cite{sn-storage03} on in-network query processing in a
centralized database system. In order to provide such mechanism, this work
evaluates the use of shard keys based on the geographic origin of the collected
data as cited to be one of the options when working with Provenance data in 
sensor networks \cite{sn-provenance01}. As a result, the data persistence system
component was conceived using the plug-and-play capability from the DSP 
platform without doing  any changes to the current DSP Platform, providing an
scalable solution that can adjust according to any data load increase. Finally,
the selected technology provides different data access mechanisms, which
directly addressed problems related to data access and Information Fusion 
\cite{sn-info-fusion}, providing export-import capabilities and visualization through
the availability of APIs written in different popular programming languages.

### 2009-11-11 docs/marcello-dissertation/thesis/chapter2.tex
* This revision fixes the problem related to the table of contents that was not printed.   - The package for hiper-link is responsible for the bug in the preamble.tex.
* References to the misc should be done by using \ref instead of \cite.

* Renaming a few bibliographic entries.

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@635 559edde2-a4d4-44e0-96fc-4035e40d7396

Each of the nodes can be deployed in a random way such as being dropped
from an airplane, or in a manual fashion at deliberately chosen geographical 
points \cite{snintro01}. Moreover, the deployment can be done as a one-time
event or an iterative way.

However, nodes can be able to move around the deployed area or be static in
this place. In the case of the degree of the former behaviour, a sensor node
can move occasionally or continuously, being a result of an incidental 
side-effect or a desired property of the device, where they are an automotive
device or attached to another moving object. For this reason, \cite{snintro01} 
describes those types of mobility as active or passive.

@@

Such sensor nodes can be of different sizes, ranging from microscopically small
particles, small devices to bigger ones with a few feet cubic. In addition to
being limited in size, they are also limited in resources and computational
power, forcing their conception limited to have its energy stored in either
batteries or scavenged from solar cells \cite{snintro01}.

Since sensor devices range on size and capabilities, the cost for each of the
devices very from hundreds to thousands of dollars, depending on its
functionality, energy utilization, etc. For thie reason, \cite{snintro01}
explains that a sensor network may potentially contain similar sensor devices,
as they can contain dissimilar ones from each other, which characterizes the heterogeneity 
of a sensor network by the hardware capabilities such as GPS enabled or having
GSM  network connectivity.

@@

Sensor nodes can communicate in different modalities such as radio, diffuse
light, inductive or sound, which may influence on the type of protocols used to
communicate among the nodes, where they can be said to be in an organized
infrastructure or as an ad hoc way \cite{snintro01}.

@@

Last, but not least, sensor networks may be influenced its coverage designed in
its purpose, its size in terms of number of nodes and its relating lifetime. As
sensors are deployed in different areas, they can be distributed in a sparse or
dense way, yet providing necessary coverage for the environment. However, 
considering that sensor devices are very likely to constant failures, projects 
have provided redundant nodes \cite{snintro01}. This degree of coverage also 
influences on the connectivity of the sensor nodes, where they can be
considered connected when a connection is available at all times. On the other
hand, nodes can communicate with other nodes in an intermittent way or sporadic
way. In the former, the sensor node is partitioned, while in the latter the
nodes are isolated most of the time and communicates at certain moments in
time.

@@

Similar to the degree of coverage of the sensor network is the number of
sensor nodes deployed in the network, where they can be in the order of
dozens, hundreds or even thousand of nodes \cite{snintro01}.

The application of the sensor network determines how long the sensor
network may be exist. For example, the lifetime of an environmental sensor
networks \cite{snex01} may be not determined since its service serves in a
regular basis, whereas the lifetime of a sensor network to determine the number
of people in an event is limited to the event.

Finally, the design of sensor networks can be related to the constraints
related to quality-of-service (QoS) such as real-time delivery within certain
period of time or robustness such as maintaining data alive even though there
a communication link is not available \cite{snintro02}.

@@

As it is shown, the data produced by this type of sensor runs in the order of
at most hundreds of Megabytes a year, what can be considered a very low data
storage requirement. Upon collecting the data from sensors, the RTC lab staff
use automation scripts written in Matlab to process, index and distribute the
raw data in different formats, including the OPeNDAP \cite{opendap}, a format
widely used in the research institutions that promotes easier data exchange
among them.

@@

The Data Sensor Platform, in short DSP Platform, is based on a micro-kernel
architecture developed on top of the Java modular framework called OSGi
\cite{osgi}. The platform represents the foundation of the execution of
NetBEAMS, since it is executed from an embedded hardware called Gumstix
\cite{gumstix} using a Linux operating system and on top of the Java Virtual
Machine (JVM). In order to transport data from sensors hosts to the network
sink, the use of a cellular connection is used to transport the collected data
from the senso devices using the HTTP Protocol, as described by
\cite{netbeams09}. Figure \ref{fig:sf-netbeams-node} shows the
architecture of the inhanced node designed by the NetBEAMS research group.

>>>>

Each of the nodes can be deployed in a random way such as being dropped
from an airplane, or in a manual fashion at deliberately chosen geographical 
points \cite{sn-intro01}. Moreover, the deployment can be done as a one-time
event or an iterative way.

However, nodes can be able to move around the deployed area or be static in
this place. In the case of the degree of the former behaviour, a sensor node
can move occasionally or continuously, being a result of an incidental 
side-effect or a desired property of the device, where they are an automotive
device or attached to another moving object. For this reason, \cite{sn-intro01} 
describes those types of mobility as active or passive.

@@

Such sensor nodes can be of different sizes, ranging from microscopically small
particles, small devices to bigger ones with a few feet cubic. In addition to
being limited in size, they are also limited in resources and computational
power, forcing their conception limited to have its energy stored in either
batteries or scavenged from solar cells \cite{sn-intro01}.

Since sensor devices range on size and capabilities, the cost for each of the
devices very from hundreds to thousands of dollars, depending on its
functionality, energy utilization, etc. For thie reason, \cite{sn-intro01}
explains that a sensor network may potentially contain similar sensor devices,
as they can contain dissimilar ones from each other, which characterizes the heterogeneity 
of a sensor network by the hardware capabilities such as GPS enabled or having
GSM  network connectivity.

@@

Sensor nodes can communicate in different modalities such as radio, diffuse
light, inductive or sound, which may influence on the type of protocols used to
communicate among the nodes, where they can be said to be in an organized
infrastructure or as an ad hoc way \cite{sn-intro01}.

@@

Last, but not least, sensor networks may be influenced its coverage designed in
its purpose, its size in terms of number of nodes and its relating lifetime. As
sensors are deployed in different areas, they can be distributed in a sparse or
dense way, yet providing necessary coverage for the environment. However, 
considering that sensor devices are very likely to constant failures, projects 
have provided redundant nodes \cite{sn-intro01}. This degree of coverage also 
influences on the connectivity of the sensor nodes, where they can be
considered connected when a connection is available at all times. On the other
hand, nodes can communicate with other nodes in an intermittent way or sporadic
way. In the former, the sensor node is partitioned, while in the latter the
nodes are isolated most of the time and communicates at certain moments in
time.

@@

Similar to the degree of coverage of the sensor network is the number of
sensor nodes deployed in the network, where they can be in the order of
dozens, hundreds or even thousand of nodes \cite{sn-intro01}.

The application of the sensor network determines how long the sensor
network may be exist. For example, the lifetime of an environmental sensor
networks \cite{sn-ex01} may be not determined since its service serves in a
regular basis, whereas the lifetime of a sensor network to determine the number
of people in an event is limited to the event.

Finally, the design of sensor networks can be related to the constraints
related to quality-of-service (QoS) such as real-time delivery within certain
period of time or robustness such as maintaining data alive even though there
a communication link is not available \cite{sn-intro02}.

@@

As it is shown, the data produced by this type of sensor runs in the order of
at most hundreds of Megabytes a year, what can be considered a very low data
storage requirement. Upon collecting the data from sensors, the RTC lab staff
use automation scripts written in Matlab to process, index and distribute the
raw data in different formats, including the OPeNDAP \ref{opendap}, a format
widely used in the research institutions that promotes easier data exchange
among them.

@@

The Data Sensor Platform, in short DSP Platform, is based on a micro-kernel
architecture developed on top of the Java modular framework called OSGi
\ref{osgi}. The platform represents the foundation of the execution of
NetBEAMS, since it is executed from an embedded hardware called Gumstix
\ref{gumstix} using a Linux operating system and on top of the Java Virtual
Machine (JVM). In order to transport data from sensors hosts to the network
sink, the use of a cellular connection is used to transport the collected data
from the senso devices using the HTTP Protocol, as described by
\cite{netbeams09}. Figure \ref{fig:sf-netbeams-node} shows the
architecture of the inhanced node designed by the NetBEAMS research group.
